---
title: "Comparison of IIT only with IIT with Mn & C"
author: "RSI PIpeline Solutions"
date: "`r format(Sys.time(), '%d %B, %Y')`" 

output: bookdown::word_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

## Introduction  {-}
The purpose of this document is to compare the yield strength predictions using Instrumented Indentation Testing (IIT) by itself with IIT combined with composition elements as additional predictors. IIT works by making repeated measurements of the force and displacement of a indenter on the surface of the pipe to arrive at a stress - strain relationship for the steel and calculate a yield and ultimate strength. But as with any physical measurement there are uncertainties involved that lead to a variation in predictions. It is known that certain elements have influence on the tensile strength of steel it was hypothesized that the inclusion of these elements along with the IIT might help reduce the variance seen in some IIT measurements.

## IIT Only Model {-}
The initial step this process was to review the IIT measurements relative to the destructive yield strength (YS). Although both the IIT and destructive yield strength have uncertainties, the destructive yield strength is considered the "truth" for all the analysis since the uncertainty is much less than the IIT. The plot of destructive YS vs. IIT is shown in Figure \@ref(fig:iit-only). The best fit line of the observations appear to have a slight bias to under-predicting as witnessed by the parallel offset to the 1:1 line.  The amount of under-prediction is more prominent in the higher YS values.

```{r data_load}
# compostion data ---------------------------------------------------------
library(readxl)
library(tidyverse)
library(tidymodels)
library(performance)
library(gtsummary)
library(patchwork)

theme_set(theme_bw(16, "serif"))

comp <-
  read_excel("~/RSI/ChemistryGrade/MasterDB-SQL-2020-09-24.xlsx",
             sheet = "Composition") %>%
  janitor::clean_names() %>%
  filter(skip==FALSE) %>%
  select(group, feature, mn,c) %>%
  group_by(group, feature) %>%
  mutate(across(.cols = mn:c, as.numeric)) %>%
  summarise(across(.cols = mn:c, ~mean(.x, na.rm=T))) %>% 
  ungroup()

# IIT data ----------------------------------------------------------------

ndt <-
  read_excel("~/RSI/ChemistryGrade/MasterDB-SQL-2020-09-24.xlsx",
             sheet = "NDT") %>%
  janitor::clean_names() %>%
  rename(group = group_name) %>%
  filter(
    skip == FALSE,
    group != "Calibration Blocks",
    str_detect("Old",
               negate = T,
               vendor_test_area_label),
    reader_method == "DPT",
    vendor %in% c('ATS', 'TDW')
  ) %>%
  group_by(group, feature) %>%
  summarise(
    iitys_5 = mean(x0_5_percent_eul_ys_ksi, na.rm = T),
    iituts = mean(uts_ksi, na.rm = T),
    nsamples = n()
  ) %>%
  filter(nsamples >= 10) %>%
  ungroup()


# Tensile data ------------------------------------------------------------


tensile <-
  read_excel("~/RSI/ChemistryGrade/MasterDB-SQL-2020-09-24.xlsx",
             sheet = "Tensile") %>%
  janitor::clean_names() %>%
  filter(skip == FALSE ,
         type != "Weld" ,
         type != "GirthWeldStrip" ,
         group != "Calibration Blocks") %>%
  rename(ys5 = x0_5_percent_eul_ys_ksi,
         uts = uts_ksi) %>%
  group_by(group, feature) %>%
  summarise(ten_ys = mean(ys5, na.rm = T),
            ten_uts = mean(uts, na.rm = T)) %>%
  ungroup()


## Combine the data -----------------------------------------------------
ndt_ten_comp <- ndt %>%
  full_join(comp,
            by = c("group", "feature")) %>%
  full_join(tensile,
            by = c("group", "feature")) %>%
  drop_na()


```

```{r iit-only, warning=TRUE, message=TRUE}

ndt_ten_comp %>%
  # pivot_longer(cols = c(.fitted_iit, .fitted_comp)) %>%
  ggplot(aes(y = ten_ys, x = iitys_5)) +
  geom_jitter(
    width = 0.1,
    height = 0.1,
    alpha = 0.65,
    col = 'midnightblue'
  ) +
  coord_obs_pred() +
  geom_abline(lty = 2, col = 'grey50', lwd = 0.7) +
  geom_smooth(method = "lm", se = F) +
  scale_color_brewer(type = "div", palette = "Set1") +
  labs(title = "Observed vs. IIT Measurement",
       x = "IIT Measured YS (ksi)",
       y = "Destructive YS (ksi)")

```
  
After the initial exploration of the IIT values vs YS, a regression model was fit to the data and the results plotted in Figure \@ref(fig:iit-regress). Although the regression line follows the 1:1 line there is still significant scatter to the data. Although the two plots are similar in appearance there are differences, Figure \@ref(fig:iit-only) is the actual IIT measurements and Figure \@ref(fig:iit-regress) is the predictions based on the regression analysis.  
  
```{r iit_only model, message=FALSE}

#define the model type and engine for the analysis
lm_model <-
  linear_reg() %>%
  set_mode("regression") %>%
  set_engine("lm")

ys_iit_rec <- recipe(ten_ys ~ iitys_5 + mn + c+s ,
                      data = ndt_ten_comp) %>%
  step_center(all_predictors()) %>%
  # step_scale(all_predictors()) %>%
  step_interact(terms = ~ mn:c ) %>%
  prep() #retain data =TRUE is default


ys_bake <- bake(ys_iit_rec, new_data = NULL)

# fit the model
fitted_iit <-  fit(lm_model,
                formula = ten_ys ~ iitys_5 ,
                data = ys_bake)

# fitted_iit$fit# fitted model object

#extract coefficients
tidy_iit <- tidy(fitted_iit, conf.int=T) %>% 
  mutate(model ="IIT") 

#extract performance metrics
iit_glance <- glance(fitted_iit$fit) %>%
  mutate(model ="IIT") %>% 
  relocate(model)

#attach the preidcted values to the original data
ys_aug <- augment(fitted_iit$fit,interval = "prediction") %>%
  mutate(model="IIT")


```
 
```{r iit-regress}
ys_aug %>%
  # pivot_longer(cols = c(.fitted_iit, .fitted_comp)) %>%
  ggplot(aes(y = ten_ys, x = .fitted)) +
  geom_jitter(width = 0.1, 
              height = 0.1,
              alpha = 0.65, 
              col = 'midnightblue') +
  coord_obs_pred() +
  geom_abline(lty = 2, col = 'grey50') +
  geom_smooth(method = "lm", se = T, col='orange') +
  scale_color_brewer(type = "div", palette = "Set1") +
  labs(title = "IIT Only Model Observed vs. Predicted",
       x = "Predicted Tensile (ksi)",
       y = "Observed Tensile (ksi)")

```

The model predicts an intercept of `r signif(tidy_iit$estimate[1],3)` and a slope of `r signif(tidy_iit$estimate[2],3)`. A slope of nearly 1 does not imply that the model is a perfect predictor as it does not account for the amount of scatter in the predictions.  The $R^2$ value was `r signif(iit_glance$r.squared,3)` Which indicates a moderately strong correlation between the predicted values and the observed values but a fair amount of scatter still exists in the predictions. The observed vs.predicted for this model are shown in the next plot. The regression line appears to fall on the unity line but in reality the offset is imperceptible at this scale with an intercept of about 2 and a slope of about 1.  
  
The model has significant residuals at the higher predicted values. At a predicted yield of 70 ksi the confidence interval is between 57 and 83 ksi. This implies that there are other influences on the yield strength that are not included in the model and the $R^2$ value indicates that greater than 30% of the variance isn't accounted for by the model. This led to the investigation of other predictors that might be of use.  

## Element Investigation  {-}
In order for other predictors to be considered for augmenting the IIT model it needed to be correlated with the samples that the IIT measurement were taken from. The pool of data available for this is the element composition. Since certain elements are known to influence the YS of steel, the wt percent of several elements were used as predictors in addition to the IIT measurements. For consideration as a candidate for modeling, the element needed to be one that can be reliably be measured with NDE methods and be available throughout the range of pipeline steel grades and vintages. Some elements have significant impact on tensile strength but are difficult to measure with NDE or are only typically used in certain grades. An example of this would be Titanium which has a significant influence on YS but is typically only used in more modern high-grade steels. An example of other elements considered but eventually rejected was Phosphorus which has known effects but are difficult to measure accurately with NDE. 
  
After several exploratory models it was decided that Manganese (Mn) and Carbon (C) had the characteristics of affecting the YS of steel and can be reliably measured with NDE methods. Figure \@ref(fig:residual-mnc) shows the model residuals (predicted minus observed) for the IIT only model versus Mn and C. There are noticeable trends in the residuals for both elements.
  
```{r residual-mnc, fig.height=5}
mn_resid <- ys_aug %>% 
  bind_cols(mn =ndt_ten_comp$mn, c = ndt_ten_comp$c) %>% 
  ggplot(aes(mn, .resid))+
  geom_point(col='orangered2')+
  geom_smooth(method = "lm", se=F)+
  labs(title = "Residuals vs. Mn Content",
       y = "Residuals",
       x = "Mn (wt%)")
  
c_resid <- ys_aug %>% 
  bind_cols(mn =ndt_ten_comp$mn, c = ndt_ten_comp$c) %>% 
  ggplot(aes(c, .resid))+
  geom_point(col='maroon')+
  geom_smooth(method = "lm", se=F)+
  labs(title = "Residuals vs. C Content",
       y = "Residuals",
       x = "C (wt%)")
  
mn_resid + c_resid +plot_annotation(caption = "IIT Only Model")


```
  
## Element Affect on Yield Strength {-}  
The relationship between Manganese and Carbon relative to the observed yield strength will be examined next. This will plot the yield strength vs. the percent weight of Carbon and Manganese. This plot shows that additional Carbon or Manganese have opposite trends on the yield strength and therefore create what is called an interaction in a linear model meaning that the effect of one is not constant relative to the other. Because of this combined effect, the model will have an interaction term of Mn x C as well as the the contribution of Mn and C individually.  

Figure \@ref(fig:mnc-plot) shows the correlation between the two elements and yield strength but this should not be construed as causation. Carbon is used to increase ultimate tensile strength but since it does so at the expense of ductility and toughness the higher yield strength steels have lower Carbon content and use other elements to increase the strength. So it has a negative correlation to yield strength but indirectly so.  

```{r mnc-plot}
ndt_ten_comp %>%
  pivot_longer(cols = mn:c) %>%
  ggplot(aes(value, ten_ys)) +
  geom_jitter(
    aes(col = name),
    width = 0.1,
    height = 0.1,
    alpha = 0.6,
    show.legend = F
  ) +
  facet_wrap( ~ name, scales = "free_x") +
  geom_smooth(method = "lm", se = F) +
  labs(title = "Correlation of Mn & C to YS",
       x="wt% - log scale",
       y = "Destructive YS (ksi)")
  scale_x_log10()


```
  
## Model with Composition  
This next section will present the results of the model with the IIT predicted yield strength along with Mn and C as additional predictors. Since it is known that the affect on YS of Mn and C the model needs to account for this. This is done through the interaction term of Mn times C.  When you include interaction between terms it creates a multicolinearity where one or more of the predictors are correlated to other predictors. Therefore the inclusion of this interaction term will create a correlation that violates the assumptions for the linear model and can create instability of the coeficent estimates. The counteraction to this problem is to center the data. This is done by subtracting the mean of each of the predictors. This method removes the multicollinearity produced by interaction, but it has the added benefit of not changing the interpretation of the coefficients. If you subtract the mean, each coefficient continues to estimate the change in the mean response per unit increase in *X* when all other predictors are held constant and the intercept is interpreted as teh predicted value when all the independent varibales are at their average.  

```{r fitted2}
fitted2 <-  fit(lm_model,
                formula = ten_ys ~ iitys_5 + mn * c  ,
                data = ys_bake) 



```

